---
title: "R Notebook"
output: html_notebook
---

###**Diving into ARIMA **
When analyzing the relationship between a response $Y_{i}$ and a covariate 
$X_{i}$, it is customary to run a regression to obtain a parameter. However when
the response is measured at different points in time, regression is restricted 
by the autocorrelation of time. Common methods are no longer viable becuase 
observations are no longer independent and identically distributed. This 
dependence is what differs time series analysis from other regressive models. 
Even though it is constrained by the sequential frequency of time, it is time 
itself that offers a rich structure to the model. To handle this type of data, 
we rely on a method referred as time series analysis. 

This method splits into two approaches, which are not necesarilly mutually 
exclusive. On the one hand, we have 'time domain approach' which prioritizes 
lagged relationships. On the other hand, 'spectral domain apprach' prioritizes 
periodicity. Before leaning in any direction, visualization is crucial as it 
often suggests the method of analysis needed. 

Even though I will not be explicitly showing any complex numbers in my analysis, 
in order to understand the mathematics of the model you must know it well. It 
turns out that understanding time series, specifically ARIMA models often involve 
understanding polynomials over the complex numbers so make sure to review its 
arithmetic, geometric series, polynomials,cartesian and polar coordinates if you 
want to dive into the mathematical details of time series modeling. 

Now, back in track, we can consider a time series as a sequence of random variable 
indexed according to the order they were obtained in time. 
Ex: $x_{1}, x_{2}, x_{3}, \dots,x_{n}$ where $x_{1}$ denotes the value obtained 
at the first time period, $x_{2}$ at the second time period and $x_{n}$ at 
the $n$ time period. It is important to state that the same distance in time 
between 1 and 2 is found between 2 and 3 and so on. 

By definition any collection of random variables $\{X_{t}\}$indexed by $t$ is 
referred to as a stochastic process, and $\{x_{t}\}$ as a realization of the 
series. For this notebook, $t$ will be considered a discrete structure and will 
span over the integers of $t=0,\pm1, \pm2, \dots$

Lets go over some general definitions of time series:

**CDF**:

$$F(c_{1},\dots,c_{T})=P(X_{1}\le c_{1},\dots,X_{T} \le c_{T})$$

**Marginal CDF's**:
$$F_{t}(c_{t})=P(X_{t} \le c_{t})$$

**Mean function**:
$$\mu_{t} = \mu_{X_{t}}= E[X_{t}]$$

**Variance function**:
$$\sigma^{2}_{t} = E[(X_{t}-\mu_{t})^{2}]$$

**Autocovariance function**: describes the strength of the linear relationship 
between random variables X at time points $t$ and $s$. 

The covariance between 2 random variables $X$ and $Y$ is:  
$$Cov(X,Y) =  E[(X-\mu_{X})(Y-\mu_{Y})]$$
When $\{X_{t}\}$ is a time series we define the autocovariance function as

$$\gamma(t,s)=Cov(X_{t},X_{s})$$

**Autocorrelation function (ACF)**: describes the correlation, meaning extent at
which the variable $X$ at time points $s$ and $t$ fluctuates together.

$$\rho(t,s)=Cor(X_{t},X_{s}) =  \frac{\gamma(t,s)} {\sqrt{(\gamma(t,t) \gamma(s,s)}}$$

If $\rho =0$ we say that X and Y are uncorrelated, meaning there is no linear 
relationship between X and Y on average. 

Recall, if X and Y are independent they are uncorrelated but converse is not true!

Lets start with the basic time series model: white noise $\{W_t\}$. White noise 
is a random signal that has no pattern, just random variation. In other words 
all frequencies are equally represented. Mathematically, this process has a mean 
of zero ($E(X_t)=0$), has no correlation between its observations across time 
($Cor(X_t,X_s)=0$), and a finite variance ($\sigma_{X_t}^{2} < \inf$). The 
Autocovariance function is as follows:

$$\gamma(s,t)= \begin{cases} \sigma^2 \space if \space s=t
               \\ 0 \space \space if \space s \ne t \end{cases}$$
                
ACF is as follows:

$$\rho(s,t)= \begin{cases} 1\space if \space s=t
               \\ 0 \space  if \space s \ne t \end{cases}$$


If we build a series out of a collection of white noise, then we would be 
creating a moving average(MA). A moving average can be defined conceputally as a 
linear regression of the current value of the series against the previous 
observed white noise error terms. The first order moving average model MA(1) can 
be expressed as

$$x_t= \mu + W_t + \alpha W_t-1$$
and the $q^{th}$ order moving average can be defined as MA(q)

$$X_t= \mu + W_t + \alpha_1 W_{t-1} + \alpha_2 W_{t-2} + \dots + \alpha_q W_{t-q}$$

Next, lets examine Autoregressive models. This series happens when a value from 
a time seires is regress on previous values from that same time series. For 
example a first order AR(p=1) would be expressed as:

$$X_t= \beta X_{t-1} + W_t$$

The autoregressive, p, component is the autocorrelation order meaning the number 
of lags used for the model and represents how it will be using past data to make 
predictions.

So far, we have explained the AR(p) and MA(q) components of ARIMA. What brings 
these two together is Integration(I) which can be attributed to stationarity. 
Stationarity refers to a state where statistical parameters converge to a constant 
in relation to time (Nau, 2018). As Brockel describes in his second chapter, 
stationarizing removes the variability of time from the series in order to be 
able to extrapolate, otherwise the information of today will share no common 
ground with the information of tomorrow. You can think of it as the conceptual 
equivalence of making the observations identically distributed. For the purpose 
of this notebook, we'll only consider the definition of weakly stationary process 
where the mean does not depend on time and the autocovariance function and ACF 
depends on s and t only through |s-t|. For a stationary process we write 

$$\rho(h)=\rho(t,t+h) =  \frac{\gamma(h)} {\sqrt{(\gamma(0) \gamma(h)}}$$

Stationarity is often achieved by differencing the series. The parameter d 
represents the degree of differencing meaning how many times will the series be 
subtracting its current minus previous value. This can be derived logically from 
the ACF and PACF plots. ACF plot displays the correlation between a series and 
its lags, while PACF displays the correlation between a variable and its lags 
that is not explained by previous lags.

Now that we have broken down ARIMA model into its components, we can go over its 
application. When we have a mixed model, autocorrelation is analyzed and the model 
order is selected. In order to determine parameters we plot ACFS against lags to 
visualize critical lag points, and confirm using ARIMA model estimate for (p,d,q) 
which stand for autocorrelation order, degrees of differencing, and moving averages 
order respectively.

Finally the forecast accuracy may be evaluated by minimizing common metrics such 
as Mean Absolute Percent Error (MAPE) , Root Mean Square Error (RMSE), and Mean 
Scaled Absolute Error (MASE). A forecast is just as good as its uncertainty. 
(Torres, 2018) .


###**Modeling and forecastin using time series **

#### Pre-processing: variable selection, response is continuous(minutes of delay)
 
```{r pre-process, eval=TRUE}
#combine day, month, year and day of week into a single date variables
flight.data$DATE <- as.Date(with(flight.data, paste(2015, MONTH, DAY,sep = "-")),
                            "%Y-%m-%d")

#partition data
sample_train <- flight.data[DATE < "2015-10-1", c("DATE","DEPARTURE_DELAY")]
#sample size 200,000
sample_train <- sample_train[sample(nrow(sample_train), 75000), ]
sample_test <- flight.data[DATE >= "2015-10-1", c("DATE","DEPARTURE_DELAY")]
sample_test <- sample_test[sample(nrow(sample_test), 25000), ]
```

#### IDENTIFICATION  

This step involves decomposing the data and testing the stationarity of the series 
to determine if any differencing is needed. First we’ll decompose data in its time 
cycles: weeks and months. Moving averages is a technique used to trace series 
biggest peaks to biggest troughs while excluding random noise, meaning it captures 
the average movement of the series, therefore smoothing the fluctuation. Figure 
I below captures the average mobility of delays per week as seen in blue, 
the average mobility of delays per month as seen in green and whole series as 
seen in red. The fluctuation is higher from week to week than from month to month 
which will already hints the importance of lag 7. 

```{r prep, eval= TRUE}
library(forecast)
library(lubridate)
month <- month(sample_train$DATE)
week <- week(sample_train$DATE)
day <- yday(sample_train$DATE)
dayofweek <- wday(sample_train$DATE, label = TRUE, abbr = FALSE)
sample_train <- cbind(sample_train, month, week, day, dayofweek)
month_delay <- sample_train %>% group_by(month) %>% 
    summarise(ave_delay = mean(DEPARTURE_DELAY))
week_delay <- sample_train %>% group_by(week) %>% 
    summarise(ave_delay = mean(DEPARTURE_DELAY))
day_delay <- sample_train %>% group_by(day) %>% 
    summarise(ave_delay = mean(DEPARTURE_DELAY))
dayofweek_delay <- sample_train %>% group_by(dayofweek) %>% 
    summarise(ave_delay = mean(DEPARTURE_DELAY))

ggplot(dayofweek_delay, aes(x = dayofweek, y = ave_delay, color = dayofweek, 
                            size = 0.8)) +
    geom_path() +
    geom_point() +
    labs(ylab = "Minutes", xlab="Day of the week", 
         title = "Average delay per day of the week") +
    theme(legend.position = "none") +
    stat_smooth(color = "red", fill= "red" , method = "loess") 


count_ts <- ts(sample_train[, c('DEPARTURE_DELAY')])
sample_train$delay_ma <- ma(sample_train$DEPARTURE_DELAY, order = 7)
sample_train$delay_ma30 <- ma(sample_train$DEPARTURE_DELAY, order = 30)
ggplot() +
    geom_line(data = sample_train, aes(x = DATE, y = DEPARTURE_DELAY, 
                                       colour = "Counts")) +
    geom_line(data = sample_train, aes(x = DATE, y = delay_ma,   
                                       colour = "Weekly Moving Average")) +
    geom_line(data = sample_train, aes(x = DATE, y = delay_ma30, 
                                       colour = "Monthly Moving Average")) +
    ylab('Delay Count')
```

Next, we’ll decompose data into time series building blocks.  As seen in Figure 
2 below, there exists no seasonal component, meaning that fluctuations of the 
data are not linked to calendar cycles. Usually to capture seasonality, 2 or more 
years of data must be analyzed, therefore the data used in this research restricts 
us from observing any. Trend component shows us that the overall pattern of the 
series oscillates around 10 minutes of delay but has a huge range between 0 and 
60 minutes demonstrating the volatility of the series. And the remainder shows 
convergence to 0 which shows that error is normalized. This convergence already 
depicts that the series is stationary.  Stationarity refers to a state where 
statistical parameters converge to a constant in relation to time (Nau, 2018). 
As Brockel describes in his second chapter, stationarizing remove the variability 
of time from the series in order to be able to extrapolate, otherwise the 
information of today will share no common ground with the information of tomorrow. 
Augmented Dickey-Fuller (ADF) test gave us a p-value of 0.01, meaning that the 
series is already stationarized, requiring no further transformation. Essentially, 
the test adds lagged differences to the fitted regression model, hence shifting 
the series by a lag length before comparing it with itself to evaluate the 
autocorrelation function. (Brockwell & Davis, 2002, p.137-179) Lag length should 
minimize correlation of residuals therefore it is chosen by minimizing  BIC or AIC.

```{r seasonal, eval=TRUE}
#calculate seasonal component 
count_ma <- ts(na.omit(sample_train$delay_ma), frequency = 30)
decomp <- stl(count_ma, s.window = "periodic")
deseasonal_delay <- seasadj(decomp)
plot(decomp)
```

#### Parameter Estimation  

This second step is where autocorrelation is analyzed and the model order is 
selected. In order to determine parameters we plot ACFS against lags to visualize 
critical lag points, and confirm using  ARIMA model estimate for (p,d,q) which 
stand for autocorrelation order, degrees of differencing, and moving averages 
order respectively. The d parameter represents the degree of differencing meaning 
how many times will the series be subtracting its current minus previous value. 
This can be derived logically from the ACF and PACF plots. ACF plot displays the 
correlation between a series and its lags, while PACF displays the correlation 
between a variable and its lags that is not explained by previous lags. As we can 
observe in figure 3 below, ACF cuts off after lag 7, while PACF tails off with a 
an oscillation that might suggest an aunderlying pattern, suggesting the series
might be a MA(7). 

```{r acfs, eval= TRUE}
par(mfrow = c(1,2))
Acf(count_ma, main = '')
Pacf(count_ma, main = '')
```

We can start stationarizing model using order d=1, and differencing further if 
needed. Figure 4 below shows the new series at lag 1. As we can see the series 
seems to be oscillating around 0 with no visible trend, which means it is 
sufficient for the model.

```{r diff, eval=TRUE}
library(tseries)
count_d1 <- diff(deseasonal_delay, differences = 1)
plot(count_d1)
adf.test(count_d1, alternative = "stationary")

Acf(count_d1, main = 'ACF for Differenced Series')
Pacf(count_d1, main = 'PACF for Differenced Series')
```

By using built in functions in R, ARIMA provided an estimate of (p=3, d= 0, q= 2), 
contradicting our previous estimate of d=1. It’s important to know that parameter 
estimation is an iterative step, meaning you decide on a value and calculate its 
results, and then go back again to deciding another value and do the same to 
compare multiple combinations of parameters. After each combination, ACF plots 
are observed as well as residual error and AIC. At the end the best fit was 
obtained from (p=3, d= 1 and q=7) as seen in Figure 5 below. 

```{r evaluating, eval=TRUE}
fit <- auto.arima(deseasonal_delay, seasonal = FALSE)
fit
#if model fits correctly, wewould expect no significant autocorrelations present
tsdisplay(residuals(fit), lag.max = 45, main = "(3,0,2) Model Residuals")

#change parameters
fit2 = arima(deseasonal_delay, order = c(3,0,7))
tsdisplay(residuals(fit2),lag.max = 15, main = "(3,0,7) Model Residuals")
```

```{r forecast, eval=TRUE}
#now do the forecasting
fcast <- forecast(fit2, h = 49479)
plot(fcast)
```


