---
title: "EDA"
output: html_notebook
---

###**3. Exploratory Data Analysis**

Now I'll explore some of its variables.
Let me begin by installing a couple of graphing packages before hand.

```{r vis packages, eval=TRUE}
library(dplyr)
library(ggplot2)
```

**Reasons for delay:**

There are over a million flights whose reason for delays were not recorded. 
From the complete cases, it can be observed that delay for security is almost null, 
and that the most common reason is air system which involves airport operations, 
heavy traffic volume, and air traffic control. Therefore I would expect to have a 
high correlation between air system delay, and large busy airports. 

```{r plot1, eval=TRUE}
#remove CANCELLATION-REASON
flight.data<-flight.data[, -"CANCELLATION_REASON"]

x <- cbind(flight.data$SECURITY_DELAY, flight.data$AIRLINE_DELAY,
           flight.data$AIR_SYSTEM_DELAY, flight.data$LATE_AIRCRAFT_DELAY,
           flight.data$WEATHER_DELAY)
names(x) <- c("security", "airline", "air.system", "late.aircraft", "weather")
x <- as.data.frame(x)

total_delay <- numeric(5)
ave_delay <- numeric(5)
max_delay <- numeric(5)
number_delays <- numeric(5)
for(i in 1:5) { #get statistics
    y <- x %>% count(x[,i]) %>% rename(avg_count = n)
    colnames(y) <- c("delay", "freq")
    y <- y[-1,]
    total_delay[i] <- sum(y$freq * y$delay)
    ave_delay[i] <- total_delay[i] / sum(y$freq)
    max_delay[i] <- max(y$delay)
    number_delays[i] <- sum(y$freq) 
}

#create summary table
delay_reasons <- rbind(total_delay,ave_delay,max_delay, number_delays) 
colnames(delay_reasons) <- c("Security", "Airline", "Air System", "Late Aircraft",
                            "Weather")
rownames(delay_reasons) <- c("Total delayed minutes", "Average delay", 
                             "Maximum delay", "Number of flights delayed")
delay_reasons <- round(delay_reasons, 0)
delay_reasons

#pie chart
names <- c("Security", "Airline", "Air System", "Late Aircraft", "Weather")
pct <- round(delay_reasons[1,] / sum(delay_reasons[1,]) * 100, 2)
names <- paste(names, pct) # add percents to labels 
names <- paste(names,"%",sep = "") # ad % to labels 
plot1<- pie(delay_reasons[1,], labels = names, col = rainbow(length(names)),
    main = "Reasons for delay", radius = 1, cex = 0.8)

```



**Ranking of Airlines:**

```{r plot2, eval=TRUE}
#ranking of airlines
airline_abb <- c("AA", "AS", "B6", "DL", 
               "EV", "F9", "HA", "MQ", 
               "NK", "OO", "UA", "US", 
               "VX", "WN")
airline <- c("American", "Alaska", "JetBlue", "Delta", 
            "Atlantic Southeast", "Frontier", "Hawaiian", "American Eagle",
            "Spirit", "Skywest", "United", "US", 
            "Virgin America", "Southwest Airlines")

a_flight_count <- tapply(flight.data$AIRLINE, flight.data$AIRLINE, function(x)
    length(x) / nrow(flight.data) * 100)
a_mean <- tapply(flight.data$DEPARTURE_DELAY, flight.data$AIRLINE, mean)
a_sd <- tapply(flight.data$DEPARTURE_DELAY, flight.data$AIRLINE, sd)
a_max <- tapply(flight.data$DEPARTURE_DELAY, flight.data$AIRLINE, max)
a_min <- tapply(flight.data$DEPARTURE_DELAY, flight.data$AIRLINE, min)
airline_summary <- data.frame(airline_abb, airline, a_flight_count, a_max, a_min,
                             a_mean, a_sd)

names(airline_summary) <- c("Airline Abreviation", "Airline","% of Total Flights",
                           "Max Delay", "Min Delay", "Mean Delay", "Standard Dev
                           Delay")

plot2 <- ggplot(airline_summary, aes(x = reorder(airline, -airline_summary[,3]),
                                   y = airline_summary[,3], fill = airline)) + 
    geom_bar(stat = "identity") +
    labs(y = "Percentage", x = "", title =  "% of all flights among airlines")+
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    theme(legend.position = "none")
plot2

par(mfrow = c(2,1))
plot3 <- ggplot(flight.data, aes(x = AIRLINE, y = DEPARTURE_DELAY, fill = AIRLINE)) + 
    geom_boxplot(outlier.shape = NA) +
    labs(title = "Departure Delays per Airline",
         x = "Airline",
         y = "Delays (min)") +
    scale_x_discrete(labels = airline) +
    scale_y_continuous(limits = c(-30, 60)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    theme(legend.position = "none") # Remove legend

plot4 <- ggplot(flight.data, aes(AIRLINE, DEPARTURE_DELAY)) +
    geom_count(aes(col = AIRLINE, show.legend = FALSE)) +
    labs(title = "Departure Delays per Airline", x = "Airline",
         y = "Delays (min)") +
    scale_x_discrete(labels = airline) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    theme(legend.position = "none")
plot3;plot4

```

Created data frame categorizing airports based on the number of flights departing
from there. Combined it with another data set containing geographic locations of 
airports to create the Tableau interactive map below. 

```{r, eval= FALSE}
#pre-processing for mapping
airport_classification <- flight.data %>% group_by(ORIGIN_AIRPORT) %>% 
    summarise(total = sum(DEPARTURE_DELAY), flight = n())
aiport.count <- cut(airport_classification$flight, 
                                     breaks = c(0,100, 1000, 10000, 100000,
                                                500000, 1000000, 10000000), 
                                     labels = c("< 100","100 << 1,000",
                                                "1,000 << 10,000", "10,000 <<
                                                100,000","100,000 << 500,000", 
                                                "500,000 << 1,000,000",
                                                ">1,000,000"), 
                                     right = FALSE)
airport_classification <- cbind(airport_classification, aiport.count)
```

<iframe align = "center" width = "1000" height = "1000" src="https://public.tableau.com/views/FLightstableau/FlightsperAirport?:embed=y&:display_count=yes&publish=yes"/>

The tables below show the most common routes in 2015. It specifies the number of 
flights between these airports ('count'), and the average delay in these routes
('delay'). It is evident that as routes becomes more frequent the variability of
delay approximates 0, while with unusual routes the average delay is more volatile. 

```{r}
#common journeys and their mean delays
routes <- summarise(group_by(flight.data,ORIGIN_AIRPORT,DESTINATION_AIRPORT),
                  count = n(), delay = sum(DEPARTURE_DELAY) / n())
```

Most frequent routes:
```{r}
routes[order(routes$count,decreasing = T)[1:20],]
```

Least frequent routes:
```{r}
routes[order(routes$count,decreasing = F)[1:20],]
```

The following graph explore how delays relate to the scheduled time of departure.
Even though on average delays are close to 0 minutes per day, there is a 
higher probability of variability in delays between midnight and 5am. During those 
early times in the morning, delays can go up to almost 40 minutes in delay.  

```{r, warning=FALSE}
#how does delay vary with scheduled time
library(tidyverse)
delay_schedule <- flight.data %>% group_by(TIME) %>% summarise(ave = 
                  sum(DEPARTURE_DELAY) / n())

ggplot(aes(x = TIME, y = ave), data = delay_schedule) + 
    geom_line(colour="purple") +
    scale_x_continuous(labels = c("0.00" = "00:00 AM", "0.25" = "06:00 AM",
                              "0.50" = "12:00 PM", "0.75" = "06:00 PM", "1.00" =
                              "11:59 PM")) +
    labs(title = "Average delay throughout a 24-hour day", y= "Average delay
         in minutes")

```

On other hand, when analyzing week of the days, there is only about a minute 
difference in average delay, meaning that the day of the week the flight 
is scheduled has no significant effect on its delay time. 

```{r, plotting time, warning= FALSE}

delay_schedule <- flight.data %>% group_by(DAY_OF_WEEK) %>% summarise(ave = 
                  sum(DEPARTURE_DELAY) / n())

delay_schedule$DAY_OF_WEEK <- as.factor(delay_schedule$DAY_OF_WEEK)
par(mfrow=c(2,1))
ggplot(aes(x = DAY_OF_WEEK, y = ave, colour=DAY_OF_WEEK), data = delay_schedule) + 
    geom_point( size=3) +
    scale_x_discrete(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = c("SUNDAY", "MONDAY", "TUESDAY", "WEDNESDAY",
                               "THRUSDAY", "FRIDAY", "SATURDAY")) +
    labs(title = "Average delay per day of the week", y = "Average delay in minutes",
         x = "Day of the week") +
    ylim(-5, 5) +
    theme(legend.position = "none")

ggplot(flight.data, aes(x = as.factor(DAY_OF_WEEK), y = DEPARTURE_DELAY, 
                        fill = as.factor(DAY_OF_WEEK))) + 
    geom_boxplot(outlier.shape = 17) +
    labs(title = "Departure delays boxplot per day of the week",
         x = "",
         y = "Delay (min)") +
    scale_x_discrete(labels = c("SUNDAY", "MONDAY", "TUESDAY", "WEDNESDAY",
                                "THRUSDAY", "FRIDAY", "SATURDAY")) +
    scale_y_continuous(limits = c(-40, 60)) +
    theme(legend.position = "none") # Remove legend
```

